---
title: "nba_statistical analysis"
author: "Justin Huang"
date: "1/25/2019"
output: 
  html_document:
    keep_md: yes
  pdf_document: default
---
Loaded in the library 
```{r message = FALSE}
library(dplyr)
library(ggplot2)
library(ggrepel)
library(directlabels)
library(gridExtra)
options(max.print = 999999999)
options(scipen=12)
```

Loaded in the data
```{r echo = FALSE}
#set directory 
getwd()
setwd("/users/justinvhuang/desktop/nba_stat_salaries")
getwd()

#loading in the data
data<-read.csv("2000_2018_nba.csv")
data <- data[,-1]
``` 

##Stastical Analysis 

### Top College schools in America and where are NBA teams placing their money over the last 19 years. 

```{r echo = FALSE}

school_tally <- data %>% group_by(school) %>% 
  summarise(count_school = n(), salary =sum(as.numeric(salary)))
school_tally %>% arrange(desc(count_school)) %>% head(n =11)
top_pay_school <- school_tally %>% arrange(desc(salary)) %>% head(n =11)
top_pay_school <- top_pay_school[-1,]
ggplot(top_pay_school, aes(school,salary)) + 
  geom_bar(stat = "identity", fill = rainbow(n=length(top_pay_school$school))) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Top Paying Schools") +
  scale_y_continuous(breaks=c(0,250000000,500000000,750000000,1000000000,1250000000))
```

The none rows contain international players and players who did not go to college. 

Let us now explore what position was drafted the most.  

```{r echo = FALSE}
#Lets see the top 3 Colleges what positions they provide the most
top3colleges<-data %>% 
  filter(school == "Kentucky"|school == "Duke"|school =="North Carolina") %>% 
  group_by(Pos,height_cm,salary,name,school)
#which position got drafted the most 
top3PosTally <- top3colleges %>% 
  group_by(Pos) %>% 
  summarize(count_pos = n())
mostPos<- top3PosTally %>% arrange(desc(count_pos)) %>% head(n =5)
mostPos
```

The powerforward position was drafted the most from these schools. The most notable and recent one is Anthony Davis. 

Lets now take a look at which team in the NBA drafted the most from these 3 schools 

```{r echo =FALSE}
#which team drafted the most from these colleges 
top3TeamTally <- top3colleges %>% 
  group_by(team) %>% 
  summarize(count_team = n())
mostpicks<-top3TeamTally %>% arrange(desc(count_team)) %>% head(n=10)
mostpicks
```

Looks like Charlotte and the LA clippers drafted the most from these 3 schools. 

Taking a look at the top paid players from these schools 

```{r echo = FALSE}
#top paid players from these schools 
lotpicks<-top3colleges %>% 
  filter(DraftRound ==1) %>% 
  group_by(DraftRound,DraftNum,name,salary,school) %>% 
  summarize(count_draftround = n())
schoolpick <-lotpicks %>% arrange(desc(count_draftround)) %>% head(n = 10)
lotpickpaid<-lotpicks %>% arrange(desc(salary)) %>%  head(n = 10)
lotpickpaid
```

Anthony Davis, Harrison Barnes and JJ Redick are the top paid players.  While Davis is a superstar player the other two's salaries can be explained by the bump in salary cap and new TV deals coming in the NBA. 

###Looking at trends in the data

```{r results = FALSE}
cor(data[, sapply(data, is.numeric)],
    use = "complete.obs", method = "pearson")
```

Lets look at some obvious correlations that should make sense with NBA players 

```{r echo = FALSE}
#intresting correlations to look at that stand out 
#height and weight

ggplot(data, aes(height_cm,weight_lb)) + 
  geom_point() +
  ggtitle("NBA Player Height and Weight")
cor(data$height_cm,data$weight_lb)
#PlusMinus TeamWins

ggplot(data, aes(PlusMinus,TeamWins)) +
  geom_point() +
  ggtitle("PlusMinus of NBA Players vs TeamWins")
cor(data$PlusMinus, data$TeamWins)
```

So Height and Weight have a positive correlation. PlusMinus is stuck in the middle with some below average players stuck on good teams that win a lot of games as a whole.  Lets look at that outlier of a 60 win team that has a plus minus of more than -20.

```{r echo =FALSE}
badBigwins <- data %>% subset(TeamWins >= 60 & PlusMinus <= -30, select = c(name,team,salary,TeamWins,PlusMinus,PlayerYear))
badBigwins
```

Huge Negative plus minuses for these two players on a winning team. They could be playing on garbage time minutes when the game is already decided.

Increased minutes should lead to increased scoring. 

```{r echo = FALSE}
ggplot(data, aes(MinGames,three)) +
  geom_point() +
  ggtitle("Minutes Played vs Threes")
cor(data$MinGames,data$three)
```

Lets look at the player who made more than 5 3's a game from the plot above. 

```{r echo= FALSE}
fivethrees<-data %>% filter(three >=5)#When stephen curry broke the 3 point record for 3's in a season
fivethrees
```

This player is Stephen Curry when he broke the 3 point record by a huge margin in one season

```{r echo = FALSE}
ggplot(data,aes(MinGames,two)) +
geom_point()+
ggtitle("Minutes Played vs Twos Made")
cor(data$MinGames,data$two)
```

There is more of a positive linear correlation in twos than there are for 3's made.  Perhaps there are role players who specialize in 3 point shooting and play minutes sparringly. 

```{r echo = FALSE}
#age and salary

ggplot(data,aes(Age,salary)) + 
  geom_point() +
  ggtitle("Age vs Salary")
cor(data$Age,data$salary)
```

We can see the maximum earning period for a NBA player is around his early 30's.  


Lets look at an interesting finding.

```{r echo = FALSE}
#To and salary
ggplot(data, aes(TO, salary)) + 
  geom_point() +
  ggtitle("TO vs Salary")
cor(data$TO,data$salary)
#ppg and salary
ggplot(data, aes(ppg,salary)) +
  geom_point() +
  ggtitle("PPG vs Salary")
cor(data$ppg, data$salary)
#FG and salary
ggplot(data,aes(FG,salary)) +
  geom_point() +
  ggtitle("FG vs Salary")
cor(data$FG,data$salary)
```

The more FG and PPG a player has should lead to an increase in salary which shows a high correlation.  However, the higher the turnovers also lead to a higher salary.  Let us explore further.

```{r echo = FALSE}
#TO mins per game
ggplot(data, aes(TO, MinGames)) +
  geom_point()+
  ggtitle("TO vs Mins Played")
cor(data$TO,data$MinGames)
#FG TO top scorers make more turnovers since they hold the ball more often 
ggplot(data, aes(FG, TO)) +
  geom_point() +
  ggtitle("FG vs TO")
cor(data$FG, data$TO)
```

We can see players who play more probably Star players who handle the ball more will also have a higher chance of turning the ball over and thus there is a positive correlation between turnover and salary.  

Salary has increased due to the increase in salary cap.

```{r echo = FALSE}
#Player Year Team Salary

ggplot(data, aes(PlayerYear, tmsalary)) +
  geom_point() +
  ggtitle("Player Year vs Team Salary")
cor(data$PlayerYear, data$tmsalary)

#TeamSal and Salcap 

ggplot(data,aes(SalCap, tmsalary)) + 
  geom_point() +
  ggtitle("Team salary vs Salary Cap")
cor(data$tmsalary, data$SalCap)
#player year and sal cap


data %>% ggplot(aes(PlayerYear, SalCap)) + 
  geom_point() +
  ggtitle("Player Year vs Salary Cap")
cor(data$PlayerYear, data$SalCap)
```

The amount of money in the league has been increasing in the league.  This cannot be attributed to just inflation.  This has to be due to the 4 CBAs and the new money coming in from TV deals and the spread of social influence through social media and international expansion. 


Looking at the trend of Twos vs Threes

```{r echo = FALSE}
data %>% ggplot(aes(PlayerYear)) +
  geom_smooth(aes(y= two_att), colour = "red") +
  geom_smooth(aes(y= three_att), colour = "green")
```

There seems to be a trend of more threes than twos as the years progress. 

###Boxplot and Histogram of NBA data 

```{r echo = FALSE}
##Three point shots
ggplot(data, aes(x = cut(three, breaks = 5), y = salary)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Salary") +
  xlab("Threes Made") +
  ggtitle("Salary and Threes")

```

Tend to get a higher mean salary for more threes made.

```{r echo = FALSE}
##PPG
ggplot(data, aes(x = cut(ppg, breaks = 10), y = salary)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Salary") +
  xlab("ppg") +
  ggtitle("PPG and Salary")
```

Higher mean salary for more points scored.  

```{r echo = FALSE}
##EFG
ggplot(data, aes(x = cut(EFG, breaks = 5), y = salary)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Salary") +
  xlab("EFG") + 
  ggtitle("EFG and Salary")
```

Mean EFG for salary does not seem to increase.  More lone spread out points above for salary for star players perhaps. 

```{r echo = FALSE}
##Total Rebounds
ggplot(data, aes(x = cut(t_reb, breaks = 5), y = salary)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Salary") +
  xlab("total rebounds") +
  ggtitle("Rebound and Salary")
```

Mean salary increases with more rebounds.  However, small players who aren't getting many rebounds are getting more of a salary as well. 

```{r echo = FALSE}
## PlayerYear
ggplot(data, aes(x = cut(PlayerYear, breaks = 19), y = salary)) + 
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("salary") +
  xlab("Player Year")+
  ggtitle("PlayerYear and Salary")
```

Salary seems to increase more during the 2016, 2017 and 2018 season.  

```{r echo = FALSE}
##ALL NBA TEAM
data %>% 
  filter(nba1 =="YES" | nba2 =="YES"| nba3 == "YES") %>% 
  ggplot(aes(salary, fill =..x..)) +
  geom_histogram(col = "black", bins = 10) +
  scale_fill_gradient(low='blue', high='yellow') +
  labs(fill = "salary")+
  scale_x_continuous(breaks = round(seq(min(data$salary), max(data$salary), by = 5000000),1)) +
  ggtitle("All NBA team and Salary")
##All Defensive Team 
data %>% 
  filter(def1 =="YES" | def2 =="YES") %>% 
  ggplot(aes(salary, fill =..x..)) +
  geom_histogram(col = "black", bins = 10) +
  scale_fill_gradient(low='blue', high='yellow') +
  labs(fill = "salary")+
  scale_x_continuous(breaks = round(seq(min(data$salary), max(data$salary), by = 5000000),1)) +
  ggtitle("Defensive Team and Salary")

data %>% 
  filter(def1 =="YES" | def2 =="YES") %>% 
  group_by(salary, PlayerYear) %>% 
  summarize(name = name[which.min(salary)])
#MVP
data %>% 
  filter(mvp =="YES") %>% 
  ggplot(aes(salary, fill =..x..)) +
  geom_histogram(col = "black", bins = 8) +
  scale_fill_gradient(low='blue', high='yellow') +
  labs(fill = "salary")+
  scale_x_continuous(breaks = round(seq(min(data$salary), max(data$salary), by = 5000000),1)) +
  ggtitle("MVP and Salary")
#DPOY
data %>% 
  filter(dpoy =="YES") %>% 
  ggplot(aes(salary, fill =..x..)) +
  geom_histogram(col = "black", bins = 8) +
  scale_fill_gradient(low='blue', high='yellow') +
  labs(fill = "salary")+
  scale_x_continuous(breaks = round(seq(min(data$salary), max(data$salary), by = 5000000),1)) +
  ggtitle("DPOY and Salary")
#All Rookie Team
data %>% 
  filter(rook1 =="YES" | rook2 =="YES") %>% 
  ggplot(aes(salary, fill =..x..)) +
  geom_histogram(col = "black", bins = 8) +
  scale_fill_gradient(low='blue', high='yellow') +
  labs(fill = "salary")+
  scale_x_continuous(breaks = round(seq(min(data$salary), max(data$salary), by = 1000000),1)) +
  ggtitle("Rookie Awards and Salary") 
```

Defensive player of the year seems to be more concentrated in the center.  The other Award categories are more spread out but still have more players in the center rather at the ends of the histogram. 

###density plot of the data

```{r echo = FALSE}
#three distribution
ggplot(data, aes(x = three)) + 
  geom_density(col = 'black', fill = 'red') +
  ggtitle("Spread of Threes")
```

Threes are right skewed as expected since the best players take and make the most shots. 

```{r echo = FALSE}
#PlusMinus distribution
ggplot(data, aes(x = PlusMinus)) + 
  geom_density(col = 'black', fill = 'red') +
  ggtitle("PlusMinus Distribution")
```

PlusMinus is more normal for the distribution. 

```{r echo = FALSE}
#Twos distribution
ggplot(data, aes(x = two)) + 
  geom_density(col = 'black', fill = 'red') +
  ggtitle("Two point distribution")
```

Two point scored is right skewed as well but not as right skewed as threepoint shots.  It would be interesting to split the range of the two point shot from 2 ft and 15ft etc...

```{r echo = FALSE}
#salary
ggplot(data, aes(x = salary)) + 
  geom_density(col = 'black', fill = 'green') +
  ggtitle("Salary distribution")
```

Salary distribution is right skewed as expected since the best players take the most money from the team.  This is due to how the contracts are strucured where players who sign for the max take 25 percent to 40 percent of the teams salary. 

```{r echo = FALSE}
#total rebounds
ggplot(data, aes(x = t_reb)) + 
  geom_density(col = 'black', fill = 'blue') +
  ggtitle("Total rebound distribution")
```
```{r echo = FALSE}
#points per game
ggplot(data, aes(x = ppg)) + 
  geom_density(col = 'black', fill = 'pink') +
  ggtitle("Point Distribution")
```

Total rebounds and PPG are also right skewed as more talented players or top players will score or grab more rebounds. 

```{r echo = FALSE}
#EFG
ggplot(data, aes(x = EFG)) + 
  geom_density(col = 'black', fill = 'yellow') +
  ggtitle("EFG distribution")
```

EFG looks to be normally distributed as it calculateds the effective field goal percentage for scoring stats.

```{r echo = FALSE}
#All rookie team 
data %>% filter(rook1 =="YES" | rook2 =="YES") %>% 
  ggplot(aes(x = salary)) + 
  geom_density(col = 'black', fill = 'yellow') +
  ggtitle("Rookie Salary Distribution")
#All NBA team
data %>% 
  filter(nba1 =="YES" | nba2 =="YES"| nba3 == "YES") %>% 
  ggplot(aes(x = salary)) + 
  geom_density(col = 'black', fill = 'yellow') +
  ggtitle("ALL NBA SALARY DISTRIBUTION")
#All Defense team
data %>% 
  filter(def1 =="YES" | def2 =="YES") %>% 
  ggplot(aes(x = salary)) + 
  geom_density(col = 'black', fill = 'yellow') +
  ggtitle("All Defensive team salary distribution")
```

The distribution for defensive player of the year looks interesting as there seems to be a second hump forming for the density plot.

```{r echo = FALSE}
#salary for internationals
data %>% 
  filter(country != "USA") %>% 
  ggplot(aes(x = salary)) + 
  geom_density(col = 'black', fill = 'blue') +
  ggtitle("International Salary Distributions")
#salary for highschoolers 
data %>% 
  filter(country == "USA" & school == "None") %>% 
  ggplot(aes(x = salary)) + 
  geom_density(col = 'black', fill = 'blue') +
  ggtitle("High School Salary Distribution")
#salary for first round draft picks 
data %>% 
  filter(DraftNum == 1) %>% 
  ggplot(aes(x = salary)) + 
  geom_density(col = 'black', fill = 'blue') +
  ggtitle("Number one draft pick Salary Distribution")
#salary for lottery picks 
data %>% 
  filter(DraftRound == 1) %>% 
  ggplot(aes(x = salary)) + 
  geom_density(col = 'black', fill = 'blue') +
  ggtitle("Round 1 Salary distribution")

data %>% 
  filter(DraftRound == 2) %>% 
  ggplot(aes(x = salary)) + 
  geom_density(col = 'black', fill = 'blue') +
  ggtitle("Round 2 salary Distribution")
```


All the plots are right skewed.  Round 2 is the most right skewed as its hard to find really gems in the second round who will be star players and command a higher salary. 